\CRANpkg{lg}: An R package for Local Gaussian Approximations

### 2021-02-17 submit/otneim.tex
Resubmitted to The R Journal

\begin{figure}[t]
    \centering
    \begin{subfigure}[t]{0.48\textwidth}
      \includegraphics[width=\textwidth]{gaussian-example}
      \caption{Estimated local Gaussian correlations between two  jointly normal variables having correlation equal to 0.5, based on 1000 observations.}
    \end{subfigure}
    \begin{subfigure}[t]{0.48\textwidth}
      \includegraphics[width=\textwidth]{returns-example}
      \caption{Local Gaussian correlation between daily returns from the CAC40 and FTSE100 stock indices, based on 1000 consecutive trading days.}
    \end{subfigure}
    \caption{Two dependence maps}
    \label{fig:initial-example}
  \end{figure}

@@

Consider next the stationary time series $\{X_t\}$. The autocorrelation function (ACF) $\rho_k = \rho(X_t, X_{t-k})$ is a well known concept for describing the serial dependence in the time series, but the ACF is, again, only capable to completely capture \emph{linear} serial dependence. \citet{laca:tjos:2017} seek to remedy this by rather calculating the the local correlation between $X_t$ and $X_{t-k}$. This leads to a test for serial independence in the natural way. In fact, this work is mainly a theoretical exercise in order to accommodate time series dependence. Testing for independence between $X_t$ and $X_{t-k}$ using observations $\{X_t, X_{t-k}\}_{t = k+1}^T$ leads to the same test statistic \eqref{eq:test-statistic1} and bootstrap procedure as the test for independence between $X_1$ and $X_2$ that we described above.

@@

We find another application of the local Gaussian approximation in the work by \citet{stov:tjos:huft:2014}, who measure and test for financial contagion. They define contagion by "a significant increase in cross-market linkages after a shock to one country" \citep{forb:rigo:2002} and employ the LGC to quantify this potential linkage. The authors estimate the LGC on a grid $\{x_1, x_2\}_k$, $k=1,\ldots,K$ along the diagonal $D=\{(x_1,x_2): x_1 = x_2\}$ \emph{before} and \emph{after} some critical event in the financial markets, denoted as the crisis (C) and the non-crisis (NC) periods respectively. They compare the two estimates using the following test statistic;

@@

\begin{equation}
\mu_i(z_i) \stackrel{\textrm{def}}{=} 0 \,\, \textrm{ and } \sigma_i(z_i) \stackrel{\textrm{def}}{=} 1, \,\, i = 1,\ldots,p,
\label{eq:simp}
\end{equation}
so that the only parameter functions left to estimate are the pairwise local Gaussian correlations $R(z) = \{\rho_{ij}(z_i, z_j)\}_{i<j}$. We use the notation $Z$, $z_i$ and $z_j$ to signify that the estimation is performed on the (approximate) standard normal scale, or $z$-scale for short. We can then estimate the joint density $f_Z(z)$ of $Z$ as
\begin{equation}
\widehat f_Z(z) = \psi(z; \mu(z) = 0, \sigma(z) = 1, R = \widehat R(z)),
\label{eq:transformed-density}
\end{equation}
where $\mu(z) = \{\mu_i(z)\}$ and $\sigma(z) = \{\sigma_i(z)\}$ for $i=1,\ldots,p$, and then substitute $f_Z$ for $\widehat f_Z$ in the following relation obtained by \citet{otne:tjos:2017} in order to estimate the unknown density $f_X$:
\begin{equation}
f(x) = f_{Z}\big(\Phi^{-1}(F_1(x_1)), \ldots, \Phi^{-1}(F_p(x_p))\big) \times \prod_{i=1}^p \frac{f_i(x_i)}{\phi\big(\Phi^{-1}(F_i(x_i))\big)},
\label{eq:backtrans}
\end{equation}
where $\phi(\cdot)$ is the standard normal pdf. This estimator is implemented the \CRANpkg{lg}-package as demonstrated in Section \ref{chap:density}.

@@

Consider finally the work by \citet{otne:tjos:2019} who take the off-diagonal element in the local correlation matrix corresponding to the local conditional covariance matrix \eqref{eq:cond-sig} or \eqref{eq:cond-sig-trans} as a local measure of conditional dependence between two stochastic variables \(X^{(1)}\) and \(X^{(2)}\) given the stochastic vector \(X^{(3)}\). Furthermore, in the case with data having been transformed to marginal standard normality, they take the statistic
\begin{equation}
T^{CI}_{n, b} = \int h(\widehat\alpha(z)) \, \textrm{d}F_n(z)
\label{eq:condintstat}
\end{equation}
as a measure of \emph{global} conditional dependence. The \CRANpkg{lg}-package provides two key functions in this framework. The first, \code{partial\_cor()}, calculates the local partial correlations as well as their estimated standard deviations on a specified grid in the \((x^{(1)}, x^{(2)}, x^{(3)})\)-space, and is essentially a wrapper function for the \code{clg()}-function presented in Section \ref{chap:density}. See Table \ref{tab:arguments-partial-cor} for details. The second function, \code{ci\_test()}, performs a test for conditional independence between the first two variables in a data set given the remaining variables using the test statistic (\ref{eq:condintstat}) and a special bootstrap procedure (described briefly below) for approximating the null distribution.

@@

The statistical literature has seen a number of applications of local Gaussian approximations in the last decade, covering several topics in dependence modeling and inference, as well as the estimation of multivariate density and conditional density functions. In this paper, we demonstrate the implementation of these methods i the R programming language using the \CRANpkg{lg}-package, as well as the graphical representation of the estimated local Gaussian correlation. The package is complete, in the sense that all major methods that have been published within this framework is now easily accessible to the practitioner. The package is also designed with a modular infrastructure that allows future methodological developments using local Gaussian approximations to be easily added to the package.

>>>>

\begin{figure}[t]
    \centering
    \begin{subfigure}[t]{0.48\textwidth}
      \includegraphics[width=\textwidth]{gaussian-example}
      \caption{Estimated local Gaussian correlations between \\ two jointly normal variables having correlation \\ equal to 0.5, based on 1000 observations.}
    \end{subfigure}
    \begin{subfigure}[t]{0.48\textwidth}
      \includegraphics[width=\textwidth]{returns-example}
      \caption{Local Gaussian correlation between daily returns from the CAC40 and FTSE100 stock indices, based on 1000 consecutive trading days.}
    \end{subfigure}
    \caption{Two dependence maps}
    \label{fig:initial-example}
  \end{figure}

@@

Consider next the stationary time series $\{X_t\}$. The autocorrelation function (ACF) $\rho_k = \rho(X_t, X_{t-k})$ is a well known concept for describing the serial dependence in the time series, but the ACF is, again, only capable of completely capturing \emph{linear} serial dependence. \citet{laca:tjos:2017} seek to remedy this by rather calculating the the local correlation between $X_t$ and $X_{t-k}$. This leads to a test for serial independence in the natural way. In fact, this work is mainly a theoretical exercise in order to accommodate time series dependence. Testing for independence between $X_t$ and $X_{t-k}$ using observations $\{X_t, X_{t-k}\}_{t = k+1}^T$ leads to the same test statistic \eqref{eq:test-statistic1} and bootstrap procedure as the test for independence between $X_1$ and $X_2$ that we described above.

@@

We find another application of the local Gaussian approximation in the work by \citet{stov:tjos:huft:2014}, who measure and test for financial contagion. They define contagion as "a significant increase in cross-market linkages after a shock to one country" \citep{forb:rigo:2002} and employ the LGC to quantify this potential linkage. The authors estimate the LGC on a grid $\{x_1, x_2\}_k$, $k=1,\ldots,K$ along the diagonal $D=\{(x_1,x_2): x_1 = x_2\}$ \emph{before} and \emph{after} some critical event in the financial markets, denoted as the crisis (C) and the non-crisis (NC) periods respectively. They compare the two estimates using the following test statistic;

@@

\begin{equation}
\mu_i(z_i) \stackrel{\textrm{def}}{=} 0 \,\, \textrm{ and } \sigma_i(z_i) \stackrel{\textrm{def}}{=} 1, \,\, i = 1,\ldots,p,
\label{eq:simp}
\end{equation}
so that the only parameter functions left to estimate are the pairwise local Gaussian correlations $R(z) = \{\rho_{ij}(z_i, z_j)\}_{i<j}$. We use the notation $Z$, $z_i$ and $z_j$ to signify that the estimation is performed on the (approximate) standard normal scale, or $z$-scale for short. We can then estimate the joint density $f_Z(z)$ of $Z$ as
\begin{equation}
\widehat f_Z(z) = \psi(z; \mu(z) = 0, \sigma(z) = 1, R = \widehat R(z)),
\label{eq:transformed-density}
\end{equation}
where $\mu(z) = \{\mu_i(z)\}$ and $\sigma(z) = \{\sigma_i(z)\}$ for $i=1,\ldots,p$, and then substitute $f_Z$ for $\widehat f_Z$ in the following relation obtained by \citet{otne:tjos:2017} in order to estimate the unknown density $f_X$:
\begin{equation}
f(x) = f_{Z}\big(\Phi^{-1}(F_1(x_1)), \ldots, \Phi^{-1}(F_p(x_p))\big) \times \prod_{i=1}^p \frac{f_i(x_i)}{\phi\big(\Phi^{-1}(F_i(x_i))\big)},
\label{eq:backtrans}
\end{equation}
where $\phi(\cdot)$ is the standard normal pdf. This estimator is implemented in the \CRANpkg{lg}-package as demonstrated in Section \ref{chap:density}.

@@

Consider, finally, the work by \citet{otne:tjos:2019} who take the off-diagonal element in the local correlation matrix corresponding to the local conditional covariance matrix \eqref{eq:cond-sig} or \eqref{eq:cond-sig-trans} as a local measure of conditional dependence between two stochastic variables \(X^{(1)}\) and \(X^{(2)}\) given the stochastic vector \(X^{(3)}\). Furthermore, in the case with data having been transformed to marginal standard normality, they take the statistic
\begin{equation}
T^{CI}_{n, b} = \int h(\widehat\alpha(z)) \, \textrm{d}F_n(z)
\label{eq:condintstat}
\end{equation}
as a measure of \emph{global} conditional dependence. The \CRANpkg{lg}-package provides two key functions in this framework. The first, \code{partial\_cor()}, calculates the local partial correlations as well as their estimated standard deviations on a specified grid in the \((x^{(1)}, x^{(2)}, x^{(3)})\)-space, and is essentially a wrapper function for the \code{clg()}-function presented in Section \ref{chap:density}. See Table \ref{tab:arguments-partial-cor} for details. The second function, \code{ci\_test()}, performs a test for conditional independence between the first two variables in a data set given the remaining variables using the test statistic (\ref{eq:condintstat}) and a special bootstrap procedure (described briefly below) for approximating the null distribution.

@@

The statistical literature has seen a number of applications of local Gaussian approximations in the last decade, covering several topics in dependence modeling and inference, as well as the estimation of multivariate density and conditional density functions. In this paper, we demonstrate the implementation of these methods in the R programming language using the \CRANpkg{lg}-package, as well as the graphical representation of the estimated local Gaussian correlation. The package is complete, in the sense that all major methods that have been published within this framework is now easily accessible to the practitioner. The package is also designed with a modular infrastructure that allows future methodological developments using local Gaussian approximations to be easily added to the package.
